---
title: "Survival Analysis Homework 5"
author: "Caesar (Zexuan) Li"
date: "11/4/2018"
output: html_document
---

```{r, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=200), warning=FALSE)
library(survival) # loading survival functions into R
library(KMsurv) 
library(dplyr)
library(cmprsk)

```


## Problem 4.10 (a)

```{r, message=FALSE}
data(bmt)
bmt.low = bmt[which(bmt$group == 2), ]
bmt.high = bmt[which(bmt$group == 3), ]

# Make the steps from the lab a function to avoid repeated code
est.cif = function(x) {
  x$ct_time <- pmin(x$t1, x$t2)
  x$ct_status <- ifelse(x$d3 == 0, 0, 1)
  x$ct_status <- ifelse(x$d2 == 1, x$ct_status * 1, 
                              x$ct_status * 2)
  
  fit.ci <- cuminc(ftime = x$ct_time, fstatus = x$ct_status, cencode = 0)
  fit.ci.relapse <- data.frame(do.call(cbind, fit.ci$'1 1')) #Time to relapse
  fit.ci.death <- data.frame(do.call(cbind, fit.ci$'1 2')) #Time to death in remission
  
  # I believe the reason why obs were duplicated was because this cmprsk package keeps all original 
  # observations, unlike how "survival" package would delete duplicates and only keep critical points
  
  fit.ci.relapse <- fit.ci.relapse %>% group_by(time) %>% slice(n()) %>% ungroup() 
  fit.ci.death <- fit.ci.death %>% group_by(time) %>% slice(n()) %>% ungroup()
  return(list(fit.ci.death, fit.ci.relapse))
}
```

For reference I printed out first 15 estimates for reference, but estimate for cumulative incidence of death at one year is the 9th row (since it's a step function), which is >288 days and < 390 days

```{r}
fit.ci = est.cif(bmt.low)
head(fit.ci[[1]], 15)
```
Therefore the answer is 0.167 for death cumulative incidence at one year

```{r}
head(fit.ci[[2]], 10)
```

For relapse, the estimated cumulative incidence is 0.074

Now for High Risk Group, risk being death

```{r}
fit.ci3 = est.cif(bmt.high)
head(fit.ci3[[1]], 15)
```

the estimate is 0.267

For risk being relapse,
```{r}
head(fit.ci3[[2]], 20)
```


The estimate is 0.356


## Problem 4.10 (b)

Referring to the results produced in part (a), low risk group, death cumulative incidence at one year, $SE=\sqrt{0.00239}=0.049$    
low risk, relapse cumulative incidence at one year $SE=\sqrt{0.0013}=0.036$    

In high risk group, death cumulative incidence at one year, $SE=\sqrt{0.00451}=0.0672$    
high risk group, relapse cumulative incidence at one year, $SE=\sqrt{0.00527}=0.0726$


## Problem 8.1 (a)

if NHL Allo is the reference hazard, then if the patient is in this group, code a variable Allo_Auto = 0 and NHL_HOD = 0. If a patient is in the NHL Auto group, then code Allo_Auto = 1 and NHL_HOD = 0. Similarly, a patient in HOD Allo group would have Allo_Auto = 0 and NHL_HOD = 1. A patient in in HOD Auto group would have Allo_Auto = 1 and NHL_HOD = 1.

## Problem 8.1 (b)

Have the same variables in (a) for main effects, namely, Allo_Auto and NHL_HOD. On top of the main effects, create an interaction variable, trans_disease, and give it values trans_disease = Allo_Auto x NHL_HOD (product of the two). Thus trans_disease would be equal to one only if a patient has both HOD and receives autologous transplant.

## Problem 8.1 (c)

If we name Allo_Auto Z1, NHL_HOD Z2 and interaction term Z3,    
$h(t|HOD Allo) = h_0(t)exp(2)$ would, according to my dummy variables, have Z1=1, Z2=0, and Z3=0.    
$h(t|NHL Auto) = h_0(t)exp(1.5)$ would have Z1=0, Z2=1, Z3=0.    
$h(t|HOD Auto) = h_0(t)exp(.5)$ would have Z1=1, Z2=1, Z3=1.    
And if the 3 coefficients are $\beta_1, \beta_2, \beta_3$, then we have 3 equations,    
$1*\beta_1+0*\beta_2+0*\beta_3 =2$,     
$0*\beta_1+1*\beta_2+0*\beta_3 =1.5$,    
$1*\beta_1+1*\beta_2+1*\beta_3 =.5$.    
Now solve the 3 equations we get $\beta_1 =2,  \beta_2=1.5, \beta_3 =-3$

## Problem 8.2 (a)

Code Z1 = 0 if the patient received the routine bathing care, and Z1 = 1 if the patient received the new bathing solution. No variable needed for the total area burned. Therefore we would have $h(t|z)=h_0(t)exp(\beta_1Z_1)$. Our test would be to test $H_0: \beta_1=0$

## Problem 8.2 (b)

Code Z1 = 0 if the patient received the routine bathing care, and Z1 = 1 if the patient received the new bathing solution. In addition, code Z2 = total surface area burned. And Z3 = Z1*Z2, which is an interaction term. The model, $$h(t|z)=h_0(t)exp(\beta_1Z_1+\beta_2Z_2+\beta_3Z_3)$$
the test is, $\beta_1=\beta_3=0$


